---
title: "Final Report"
author: "Ben, Andrew, Pranati, Filip"
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
    number-sections: true
editor: source
execute: 
  error: true
  echo: false
  message: false
  warning: false
---
### Data Description
The data set hopes to discover a relationship between the mean household income and amount of tons in CO2 emission on an average country. The study is observational with the experimental units being each country.The mean household income data provides the average household income per year for each individual country, starting at 1950 and ending in 2018. The units of the incomes provided are in standardized international dollar units. The CO2 emissions per person dataset records the average carbon dioxide produced per person in the country per year measured in metric tonnes, starting in 1950 and ending in 2018.

### Data Cleaning
After looking at both sets of data we decided to start our analysis in 1950, as this is when the CO2 emission data began to be widespread. The CO2 data was already clean and ready for use but the mean household income data was not. To fix  this, we had to replace the "k's" in the data and multiply by 1000. After importing the selected years and cleaning the data, we pivoted the data to long format. We then full joined the two data sets together by country and year to create one large data set. This does leave NA's in the dataset for some countries where CO2 data is only recent, but we are still deciding how we want to treat those few, small countries. 

## Data
```{r}
CO2 <- read_csv("co2_emissions_tonnes_per_person.csv", col_select = c(1, 152:220))
mhi <- read_csv("mhhinc.csv", col_select = c(1, 152:220))


mhi <- mhi %>%
  mutate_at(vars(-1), function(x) {
    x <- ifelse(str_detect(x, "k"), 
                as.numeric(str_replace_all(x, "k", "")) * 1000, 
                as.numeric(x))
    return(x)
})

mhi <- pivot_longer(
  mhi, cols = 2:70, names_to = "year", values_to = "mean_income",
  values_transform = list(mean_income = as.double)
    )
CO2 <- pivot_longer(CO2, cols = 2:70, names_to = "year", values_to = "co2")

mhi_CO2 <- mhi %>% 
  full_join(CO2, by = join_by(country, year)) %>% 
  mutate_all(~replace(., is.na(.), 0))
```

## Data Visualization

:::{callout-tip}
Double clicking a country will select that country and deselect all of the other countries. Single clicking any other country will re-add that country to the plot. Double clicking any selected country will re-add all the countries to the plot.
:::

```{r}
co2_income_plot <- mhi_CO2 %>% 
  ggplot(mapping = aes(x = mean_income, y = sqrt(co2), color = country)) +
  geom_line(show.legend = FALSE, linejoin = "round") +
  labs(
    x = "Mean Household Income [International $]",
    y = "",
    title = "Carbon Dioxide Produced vs Mean Household Income"
  )

interactive_co2_income_plot <- ggplotly(co2_income_plot) %>% 
  layout(title = list(text = paste0(
    "Carbon Dioxide Produced vs Mean Household Income",
    "<br>",
    "<sup>",
    "Carbon Dioxide Produced per Person [Metric Tonnes]",
    "</sup>")))
interactive_co2_income_plot

time_plot <- mhi_CO2 %>% 
  mutate(
    co2_per_income = sqrt(co2)/mean_income
  ) %>% 
  ggplot(mapping = aes(x = year, y = co2_per_income, color = country, fill = country)) +
  geom_col(position = "dodge", show.legend = FALSE) +
  scale_x_discrete(
    breaks = c("1950", "1960", "1970", "1980", "1990", "2000", "2010", "2018")
    ) +
  labs(
    x = "Year",
    y = "",
    title = "Metric Tons of Carbon Dioxide Produced per Person by Mean Household Income per Year"
  )

interactive_time_plot <- ggplotly(time_plot) %>% 
  layout(title = list(text = paste0(
    "Carbon Dioxide Produced per Person by Mean Household Income per Year",
    "<br>",
    "<sup>",
    "Carbon Dioxide Produced per Person by Mean Household Income [Metric Tonnes / International $]",
    "</sup>")))
  
interactive_time_plot
```

The two plots shown are interactive plots of Carbon Dioxide Produced per Person vs. Mean Household Income and Carbon Dioxide Produced per Person by Mean Household Income per Year, respectively. The first plot shows Mean Household Income on the x-axis and metric tonnes of Carbon Dioxide produced per person on the y-axis. This plot shows how increasing income affects the amount of CO2 produced per person. The second plot shows the change in graph one over time, from 1950 to 2018. This plot shows how the ratio of CO2 produced over mean household income has changed over the last 68 years.

## Linear Regression

```{r}
model <- lm(sqrt(co2) ~ mean_income, data = mhi_CO2)
summary(model)
```

Linear regression involves modeling the relationship between variables in a dataset in a scalar way. A linear model in the form yi = β0 + β1xi + ε is used to describe the relationship between each explanatory variable to the response accounting for random error. In this study, mean income is the explanatory variable and predicted co2 level is the response, so the model attempts to explain how changes in mean income increase or decrease predicted(co2) levels. 

$Predicted(\sqrt{co2}) = 0.9085 + .0001402(mean income)$

For every one unit increase in mean income(measured in standardized international dollar units), we expect a .0001402 increase in the square root of co2 emission levels(measured in metric tons), or 0.3091 lb increase.

Our model has an r value of .6368, which shows a moderate, positive linear relationship between yearly CO2 emissions and mean household income. Our R\^2 value(.4057) shows that household income only accounts for 40.57% of variation in CO2 emissions. This is slightly strong, and we would want to look to manipulate our current variables or add more explanatory variables to make our model stronger. Since a model without transformations has a relatively low R^2 value of 0.33, a square root transformation has been applied to the response variable, carbon dioxide emissions, to better explain changes in mean income in the model. Through this transformation on co2 levels, the R^2 value increases to 0.406 explaining around 7% more variability in co2 emissions by mean income. 

## Model Fit

```{r}
response_variance <- var(mhi_CO2$co2, na.rm = TRUE)
fitted_vals_variance <- var(model$fitted.values)
residuals_variance <- var(model$residuals)

Datasets <- c("Response Variable", "Model Fitted Values", "Model Residuals")
Variance <- c(response_variance, fitted_vals_variance, residuals_variance)
variance_df <- data.frame(Datasets, Variance)

kable(variance_df,
      format = "html",
      caption = "<center><strong>Variance of Reponse Variable and Model</strong></center>",
      align = "c") |>
  row_spec(0, underline = TRUE, color = "black", background = "lightblue") |>
  column_spec(1, color = "black")

```

The proportion of variability in the response values account for by the regression model is calculated using R-squared with a formula of (R2 = 1 - (residual variance / variance in response value)).

R2 = 1 - (32.72974 / 48.69698) = 0.327 or 32.7%.

Since our R2 is only around 33% this suggest that the quality of our model is pretty weak and would most likely be improved by increasing the number of explanatory variables.




## Simulation

```{r}
predictions <- predict(model, newdata = mhi_CO2)

predictions <- predict(model, newdata = mhi_CO2)

# Extract the residual standard error
sigma <- sigma(model)


# Generate random errors
set.seed(123) # for reproducibility
errors <- rnorm(nrow(mhi_CO2), mean = 0, sd = sigma)

# Add random errors to predictions
predictions_with_errors <- predictions + errors

# Create a prediction dataset
prediction_data <- data.frame(
  country = mhi_CO2$country,
  year = mhi_CO2$year,
  mean_income = mhi_CO2$mean_income,
  co2 = predictions_with_errors
)

```



