---
title: "CO2 and Mean Family Income Final Report"
author: "Ben, Andrew, Pranati, Filip"
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
    number-sections: true
editor: source
execute: 
  error: true
  echo: false
  message: false
  warning: false
---
```{r setup}
library(tidyverse)
library(plotly)
library(kableExtra)
```


### Data Description
The data set hopes to discover a relationship between the mean household income and amount of tons in CO2 emission on an average country. The study is observational with the experimental units being each country.The mean household income data provides the average household income per year for each individual country, starting at 1950 and ending in 2018. The units of the incomes provided are in standardized international dollar units. The CO2 emissions per person dataset records the average carbon dioxide produced per person in the country per year measured in metric tonnes, starting in 1950 and ending in 2018.

### Data Cleaning
After looking at both sets of data we decided to start our analysis in 1950, as this is when the CO2 emission data began to be widespread. The CO2 data was already clean and ready for use but the mean household income data was not. To fix  this, we had to replace the "k's" in the data and multiply by 1000. After importing the selected years and cleaning the data, we pivoted the data to long format. We then full joined the two data sets together by country and year to create one large data set. This does leave NA's in the dataset for some countries where CO2 data is only recent, but we are still deciding how we want to treat those few, small countries. 

## Data
```{r}
CO2 <- read_csv("co2_emissions_tonnes_per_person.csv", col_select = c(1, 152:220))
mhi <- read_csv("mhhinc.csv", col_select = c(1, 152:220))


mhi <- mhi %>%
  mutate_at(vars(-1), function(x) {
    x <- ifelse(str_detect(x, "k"), 
                as.numeric(str_replace_all(x, "k", "")) * 1000, 
                as.numeric(x))
    return(x)
})

mhi <- pivot_longer(
  mhi, cols = 2:70, names_to = "year", values_to = "mean_income",
  values_transform = list(mean_income = as.double)
    )
CO2 <- pivot_longer(CO2, cols = 2:70, names_to = "year", values_to = "co2")

mhi_CO2 <- mhi %>% 
  full_join(CO2, by = join_by(country, year)) %>% 
  mutate_all(~replace(., is.na(.), 0))
```

## Data Visualization

:::{callout-tip}
Double clicking a country will select that country and deselect all of the other countries. Single clicking any other country will add that country to the plot. Double clicking any selected country will add all the countries to the plot. It is recommended that only a few countries are displayed at a time for ease of use. All of the countries are provided to give a wide spectrum of options and to compare between different levels of industrialization and modernization.
:::

```{r}
co2_income_plot <- mhi_CO2 %>% 
  ggplot(mapping = aes(x = mean_income, y = sqrt(co2), color = country)) +
  geom_line(show.legend = FALSE, linejoin = "round") +
  labs(
    x = "Mean Household Income [International $]",
    y = "",
    title = "Square Root of Carbon Dioxide Produced vs Mean Household Income"
  )

interactive_co2_income_plot <- ggplotly(co2_income_plot) %>% 
  layout(title = list(text = paste0(
    "Square Root of Carbon Dioxide Produced vs Mean Household Income",
    "<br>",
    "<sup>",
    "Square Root of Carbon Dioxide Produced per Person [Metric Tonnes]",
    "</sup>")))
interactive_co2_income_plot

time_plot <- mhi_CO2 %>% 
  mutate(
    co2_per_income = sqrt(co2)/mean_income
  ) %>% 
  ggplot(mapping = aes(x = year, y = co2_per_income, color = country, fill = country)) +
  geom_col(position = "dodge", show.legend = FALSE) +
  scale_x_discrete(
    breaks = c("1950", "1960", "1970", "1980", "1990", "2000", "2010", "2018")
    ) +
  labs(
    x = "Year",
    y = "",
    title = "Square Root of Metric Tons of Carbon Dioxide Produced per Person by Mean Household Income per Year"
  )

interactive_time_plot <- ggplotly(time_plot) %>% 
  layout(title = list(text = paste0(
    "Square Root of Carbon Dioxide Produced by Mean Household Income per Year",
    "<br>",
    "<sup>",
    "Square Root of Carbon Dioxide Produced per Person by Mean Household Income [Metric Tonnes / International $]",
    "</sup>")))
  
interactive_time_plot
```

The two plots shown are interactive plots of the square root of Carbon Dioxide Produced per Person vs. Mean Household Income and the square root of Carbon Dioxide Produced per Person by Mean Household Income per Year, respectively. The first plot shows Mean Household Income on the x-axis and the square root of metric tonnes of Carbon Dioxide produced per person on the y-axis. This plot shows how increasing income affects the amount of CO2 produced per person. The second plot shows the change in graph one over time, while using the mean income over time, from 1950 to 2018. This plot shows how the ratio of the square root of CO2 produced over mean household income has changed over the last 68 years.

Looking at the first plot shows the effect that income has on CO2 produced. It shows for most countries that as mean household income increases, so does CO2 produced. The data also shows that for industrialized countries there is a point at which income and CO2 become decoupled and CO2 per person starts to fall. We believe this is due to green technologies being more cost prohibitive and therefore there effects are mainly seen at higher income levels. As green technologies become cheaper to produce, we expect that this trend will continue and will spread to lower income people as well. 

The second plot shows how income and CO2 produced have changed over time. This graph follows a similar shape for most countries, with an early peak that is followed by an decay which is very similar in shape to $1/x$. This shows that as a country industrializes, it begins to reduce the amount of CO2 produced per person. A country that shows this effect especially well is China, whose CO2 per mean household income peaks in 1961 and has been falling since. Countries that have already industrialized stay extremely constant.

## Linear Regression

```{r}
model <- lm(sqrt(co2) ~ mean_income, data = mhi_CO2)
summary(model)
```

Linear regression involves modeling the relationship between variables in a dataset in a scalar way. A linear model in the form yi = β0 + β1xi + ε is used to describe the relationship between each explanatory variable to the response accounting for random error. In this study, mean income is the explanatory variable and predicted CO2 level is the response, so the model attempts to explain how changes in mean income increase or decrease predicted(CO2) levels. 

$Predicted(\sqrt{co2}) = 0.9085 + .0001402(mean income)$

For every one unit increase in mean income(measured in standardized international dollar units), we expect a .0001402 increase in the square root of co2 emission levels(measured in metric tons), or 0.3091 lb increase.

Our model has an r value of .6368, which shows a moderate, positive linear relationship between yearly CO2 emissions and mean household income. Our R\^2 value(.4057) shows that household income only accounts for 40.57% of variation in CO2 emissions. This is slightly strong, and we would want to look to manipulate our current variables or add more explanatory variables to make our model stronger. Since a model without transformations has a relatively low R\^2 value of 0.4057, a square root transformation has been applied to the response variable, carbon dioxide emissions, to better explain changes in mean income in the model. Through this transformation on CO2 levels, the R^2 value increases to 0.406 explaining around 7% more variability in CO2 emissions by mean income. 

## Model Fit

```{r}
response_variance <- var(mhi_CO2$co2, na.rm = TRUE)
fitted_vals_variance <- var(model$fitted.values)
residuals_variance <- var(model$residuals)

Datasets <- c("Response Variable", "Model Fitted Values", "Model Residuals")
Variance <- c(response_variance, fitted_vals_variance, residuals_variance)
variance_df <- data.frame(Datasets, Variance)

kable(variance_df,
      format = "html",
      caption = "<center><strong>Variance of Reponse Variable and Model</strong></center>",
      align = "c") |>
  row_spec(0, underline = TRUE, color = "black", background = "lightblue") |>
  column_spec(1, color = "black")

```

The proportion of variability in the response values account for by the regression model is calculated using R-squared with a formula of (R2 = 1 - (residual variance / variance in response value)).

R2 = 1 - (27.6448 / 46.5166) = 0.4057 or 40.57%.

Since our R2 is only around 33% this suggest that the quality of our model is pretty weak and would most likely be improved by increasing the number of explanatory variables.


## Simulation

```{r}
#| layout-nrow: 2
predictions <- predict(model, newdata = mhi_CO2)

predictions <- predict(model, newdata = mhi_CO2)

# Extract the residual standard error
sigma <- sigma(model)

n_sims <- 1000

prediction_datasets <- vector("list", n_sims)


# Generate random errors
set.seed(123) # for reproducibility
errors <- rnorm(nrow(mhi_CO2), mean = 0, sd = sigma)

# Add random errors to predictions
predictions_with_errors <- predictions + errors

# Create a prediction dataset
prediction_data <- data.frame(
  country = mhi_CO2$country,
  year = mhi_CO2$year,
  mean_income = mhi_CO2$mean_income,
  co2 = predictions_with_errors
)

# Plot comparing the simulated data to the observed data
observed_plot <- ggplot(data = mhi_CO2,
       mapping = aes(x = mean_income, y = sqrt(co2), color = country)) +
  geom_line(show.legend = FALSE, linejoin = "round") +
  labs(
    x = "Mean Household Income [International $]",
    y = "",
    title = "Square Root of Carbon Dioxide Produced vs Mean Household Income")

interactive_observed_plot <- ggplotly(observed_plot) %>% 
  layout(title = list(text = paste0(
    "Square Root of Carbon Dioxide Produced vs Mean Household Income [Actual]",
    "<br>",
    "<sup>",
    "Square Root of Carbon Dioxide Produced per Person [Metric Tonnes]",
    "</sup>")))
interactive_observed_plot

simulated_plot <- ggplot(data = prediction_data,
       mapping = aes(x = mean_income, y = sqrt(co2), color = country)) +
  geom_line(show.legend = FALSE, linejoin = "round") +
  labs(
    x = "Predicted Mean Household Income [International $]",
    y = "",
    title = "Carbon Dioxide Produced vs Predicted Mean Household Income")

interactive_simulated_plot <- ggplotly(simulated_plot) %>% 
  layout(title = list(text = paste0(
    "Square Root of Carbon Dioxide Produced vs Mean Household Income [Simulated]",
    "<br>",
    "<sup>",
    "Square Root of Carbon Dioxide Produced per Person [Metric Tonnes]",
    "</sup>")))
interactive_simulated_plot

for(i in seq_len(n_sims)){
  
  predictions <- predict(model, newdata = mhi_CO2)
  
  errors <- rnorm(nrow(mhi_CO2), mean = 0, sd = sigma)
  
  predictions_with_errors <- predictions + errors

  prediction_data <- data.frame(
  country = mhi_CO2$country,
  year = mhi_CO2$year,
  mean_income = predictions_with_errors,
  co2 = mhi_CO2$mean_income
  )
  
  prediction_datasets[[i]] <- prediction_data
}

rSquaredList = list()

for(i in seq_len(n_sims)) {
  temp2 <- lm(sqrt(mhi_CO2$co2) ~ prediction_datasets[[i]]$mean_income)
  rSquaredList <- append(rSquaredList, summary(temp2)$r.squared)
}
```

The actual and simulated data are very different for the square root of CO2 produced vs mean household income. We believe this is a result of the large variability in our data set. The simulated data is similar to the actual data from smaller countries, but for larger countries it becomes disjointed. 

```{r}
ggplot() + 
  geom_histogram(aes(unlist(rSquaredList)), fill = "blue", color = "lightgrey") +
  labs(title = "Observed vs Predicted R-Squared Values",
     subtitle = "Count",
     x = "R-Squared",
     y = "") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

The R\^2 values calculated from the simulated data set show that our model does not generate data similar to what was observed. We observed an R\^2 value of .4057, while our 1000 simulations provided a mean R\^2 value of approximately .135. This means that our simulated data is worse than the actual data and that it only represents about 13% of the variance in the model.
